---
title: "Bootstrapping basics"
author: "Michael Love"
format: beamer
execute:
  echo: true
---

# sampling variance

some words

# no access to full "population"

# idea: sub-sampling a sample

# three types of bootstrapping

# example: line with non-normal errors

```{r}
library(dplyr)
library(infer)
library(ggplot2)
set.seed(1)
n <- 200
x <- runif(n)
eps <- rexp(n, 5)
eps <- eps - mean(eps)
slope <- .2
y <- slope * x + eps
plot(x,y)
abline(0, slope, col="red")
dat <- data.frame(x,y)
```

# simple bootstrapping

```{r}
set.seed(1)
boots <- replicate(1000, {
  idx <- sample(n, replace=TRUE)
  coef(lm(y ~ x, data=dat[idx,]))[2]
})
hist(boots)
sd(boots)
```

# using `boot`

```{r}
library(boot)
get_slope <- function(data, idx) {
  coef(lm(y ~ x, data=data[idx,]))[2]
}
boots2 <- boot(dat, get_slope, R=1000)
boots2
plot(boots2)
```

# additional plots

```{r}
library(car)
confint(boots2)
hist(boots2)
```

# {infer} package

S, H, G, C =
Specify, Hypothesize, Generate, Calculate

```{r eval=FALSE}
set.seed(1)
perm <- dat %>% specify(y ~ x) %>%
  hypothesize(null="independence") %>%
  generate(reps=1000, type="permute") %>%
  calculate(stat="slope")
```

# bootstrapping a statistic

```{r}
set.seed(1)
boot <- dat %>% specify(y ~ x) %>%
  generate(reps=1000, type="bootstrap") %>%
  calculate(stat="slope")
visualize(boot, bins=30)
```

# confidence intervals

```{r}
ci <- get_ci(boot)
ci
obs_beta <- dat %>% 
  specify(y ~ x) %>%
  calculate(stat="slope")
```

# visualize

```{r}
visualize(boot, bins=30) +
  shade_confidence_interval(ci, alpha=.3, 
                            fill="yellow", color="orange") + 
  geom_vline(xintercept=obs_beta$stat, 
             color="red", linewidth=1)
```

# versus lm

```{r}
ci
confint(lm(y ~ x), parm="x")
```

# going further

> Bootstrapping Regression Models in R
> An Appendix to An R Companion to Applied Regression, 3rd ed.
> John Fox & Sanford Weisberg

(can find PDF online)

# Fox and Weisberg example

```{r}
library(car)
fit <- lm(y ~ x)
summary(fit)$coef
set.seed(1)
bfit <- Boot(fit, method="residual")
summary(bfit)
confint(bfit)
```
